Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-08-11 paper/manuscript.tex
Fix undefined citations

Conversely, the value of the binding free energy depends on the chosen structural definition of the bound complex. The relationship between the predictions of this theory and experimental reporters (thermochemical, spectroscopic, inhibition kinetics, and others) of complex formation has been discussed.\cite{gilson1997statisticalthermodynamic,mihailescu2004theory,gallicchio2011recentadv,simonson2016physical,gallicchio2021comppeptsci} The interpretation of experimental reporters of binding is a complex issue in itself. Still, the consensus is that this theory gives nearly equivalent results when the binding is strong and specific, and the definition of the bound complex covers the major bound poses while excluding obvious unbound configurations. In some cases it is valuable to consider specific definitions of the complex by, for example, restricting the orientation of the ligand to pinpoint the relative contribution of different binding poses.

@@

Another variant of a dual coordinate approach is implemented in NAMD \cite{jiang2019computing}. Whereas Ries et al.\  use distance restraints to keep two molecules on top of each other, corresponding atoms in NAMD are held on top of each other by constraints; in practice, this means that these coordinates are forced to be identical. AMBER's free energy implementation recently switched to a similar approach \cite{Lee_2023,York_ACSPhysChemAu_2023_v3_p478}.
While the implementation technically uses two sets of coordinates for the entire transforming region, selected atoms that have a direct atom-atom mapping are constrained to have common coordinates (sometimes referred to as a "common core"), while others that are mapped into dummy atoms and are treated with softcore potentials~\cite{Lee_JChemTheoryComput_2020_v16_p5512,Tsai_JChemTheoryComput_2023_v19_p640} have separable coordinates.  Thus the common core part of the dual topology are constrained to have the same coordinates to maximize phase space overlap, whereas the softcore region has separate coordinates.  The two topologies do not see each other, and this property enables this approach to be leveraged in the alchemical enhanced sampling (ACES) method~\cite{Lee_2023,Tsai_JChemInfModel_2024_v64_p7046,Zhang_JChemTheoryComput_2024_v20_p3935}.
%The authors refer to this approach as hybrid single-dual topology. While it technically uses two separate sets of coordinates, the constraint that corresponding atoms have identical coordinates makes it effectively a single coordinate method; hence, I consider NAMD’s approach to be a variant of Pearlman’s (single coordinate) dual topology.

@@

The practical implementation of the approaches to carry out alchemial transformations above leads to additional, often non-trivial differences within each methodology. In the description of DC approaches we already pointed out that the two sets of coordinates can be held on top of each other by constraints \cite{jiang2019computing,Lee_2023} or restraints. In the latter case, the restraints may couple the two set of coordinates directly \cite{Axelsen_1998,Ries_2022}, or the restraints may be defined relative to a common reference frame \cite{rocklin2013separated,Baumann_2023}. Finally, in ATM the two copies of the coordinates are held separate \cite{Azimi_2022}.

@@

Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed. One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation . This problem is similar to the required restraints in ABFE Calculations (see \ref{ABFErest}). Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on Boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-DCDT approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other. 
However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets. \cite{jespers2019qligfep, Rieder2022Lev}

@@

There are multiple strategies to alleviate such sampling challenges in an alchemical calculation. A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}. Another approach, known as integration over parts,\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recentadv,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate. In the integration over parts approach, restraining potentials can be used to restrict the sampling to specific macrostates,\cite{khuttan2023taming}. However, in practice, restraining potentials are sometimes omitted when the states are known to be separated by large free energy barriers and do not interconvert within the simulation time scale, such as cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}. A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation. This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from states separated by large free energy barriers) via replica exchange protocols. More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}. Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda}, and optimized alchemical enhanced sampling (ACES)~\cite{lee2023aces}  and approaches that integrate well-tempered metadynamics~\cite{barducci2008well} with alchemical methods~\cite{zhou2025zooming, khuttan2024make}, to name a few.

@@

\subsubsection{Which $\lambda$-sampling scheme will work best for my problem?}
\label{sec:sampling_schemes}
Though all alchemical simulations must sample from multiple $\vec{\lambda}$ states, different approaches can be used to achieve this. Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes. The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}). This type of scheme is currently used for some AMBER TI calculations~\cite{song2019using} and for Sire as implemented in BioSimSpace~\cite{hedges2019biosimspace}. However, if these simulations can be run simultaneously with communication between them, a simple extension allows mixing between these replicas. In this approach, the simulation at each $\vec{\lambda}$ can undergo periodic exchanges with neighboring $\vec{\lambda}$ values. This form of replica exchange, called Hamiltonian replica exchange, is based on ideas developed from Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}. With the Metropolis-Hastings acceptance criterion for exchanges, the generated ensemble of all replicas still samples from the Boltzmann distribution for each replica. This approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensional,sugita1999replicaexchange, woods2003development, jiang2010free}. The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{B}. It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltonian,hritz2007optimization},
AMBER~\cite{Lee_JChemInfModel_2020_v60_p5595}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
A specific variant of Hamiltonian replica exchange is FEP/REST2~\cite{rest2, feprest} that involves ``heating'' a small portion of the protein-ligand system to enhance the sampling within the binding pocket. It has been promoted as a great way to enhance sampling and is employed by default in Schr\"{o}dinger's FEP+ package. However, it should be noted that several studies have questioned its reliability and reported that it degrades the quality of free energy predictions as compared to those based on standard MD simulations without any enhanced sampling~\cite{wan2020fep+, bhati2022}. Therefore, a blind application of this sampling scheme is not recommended and it must be used with caution if at all. A variant of this method that overcomes some of the reported issues with FEP/REST2 is called alchemical enhanced sampling (ACES)~\cite{Lee_JChemTheoryComput_2023_v19_p472}. ACES creates localized targeted "enhanced sampling states" that are connected to the real state end-points through HREMD.  The method leverages the dual topology framework to produce a concerted “counter-diffusion” of alchemical states that leads to minimal rearrangement of the environment along the alchemical path~\cite{York_ACSPhysChemAu_2023_v3_p478}. The computational overhead of the method is negligible relative to more conventional alchemical free energy simulations with HREMD.

@@

	\item \href{https://ambermd.org/}{AMBER}, including its GPU-accelerated free energy engine~\cite{Lee_JChemTheoryComput_2017_v13_p3077,Giese_JChemTheoryComput_2018_v14_p1564,Lee_JChemInfModel_2018_v58_p2043} that supports a wide range of alchemical free energy methods~\cite{Tsai_JChemTheoryComput_2023_v19_p640,Lee_2023,Zhang_JChemTheoryComput_2024_v20_p3935}, workflows~\cite{Ganguly_JChemInfModel_2022_v62_p6069} and analysis tools~\cite{Giese_JChemInfModel_2025_v65_p5273,Case_JChemInfModel_2023_v63_p6183}. 

>>>>

Conversely, the value of the binding free energy depends on the chosen structural definition of the bound complex. The relationship between the predictions of this theory and experimental reporters (thermochemical, spectroscopic, inhibition kinetics, and others) of complex formation has been discussed.\cite{gilson1997statisticalthermodynamic,mihailescu2004theory,gallicchio2011recent,simonson2016physical,gallicchio2021comppeptsci} The interpretation of experimental reporters of binding is a complex issue in itself. Still, the consensus is that this theory gives nearly equivalent results when the binding is strong and specific, and the definition of the bound complex covers the major bound poses while excluding obvious unbound configurations. In some cases it is valuable to consider specific definitions of the complex by, for example, restricting the orientation of the ligand to pinpoint the relative contribution of different binding poses.

@@

Another variant of a dual coordinate approach is implemented in NAMD \cite{jiang2019computing}. Whereas Ries et al.\  use distance restraints to keep two molecules on top of each other, corresponding atoms in NAMD are held on top of each other by constraints; in practice, this means that these coordinates are forced to be identical. AMBER's free energy implementation recently switched to a similar approach \cite{lee2023aces,York_ACSPhysChemAu_2023_v3_p478}.
While the implementation technically uses two sets of coordinates for the entire transforming region, selected atoms that have a direct atom-atom mapping are constrained to have common coordinates (sometimes referred to as a "common core"), while others that are mapped into dummy atoms and are treated with softcore potentials~\cite{Lee_JChemTheoryComput_2020_v16_p5512,Tsai_JChemTheoryComput_2023_v19_p640} have separable coordinates.  Thus the common core part of the dual topology are constrained to have the same coordinates to maximize phase space overlap, whereas the softcore region has separate coordinates.  The two topologies do not see each other, and this property enables this approach to be leveraged in the alchemical enhanced sampling (ACES) method~\cite{lee2023aces,Tsai_JChemInfModel_2024_v64_p7046,Zhang_JChemTheoryComput_2024_v20_p3935}.
%The authors refer to this approach as hybrid single-dual topology. While it technically uses two separate sets of coordinates, the constraint that corresponding atoms have identical coordinates makes it effectively a single coordinate method; hence, I consider NAMD’s approach to be a variant of Pearlman’s (single coordinate) dual topology.

@@

The practical implementation of the approaches to carry out alchemial transformations above leads to additional, often non-trivial differences within each methodology. In the description of DC approaches we already pointed out that the two sets of coordinates can be held on top of each other by constraints \cite{jiang2019computing,lee2023aces} or restraints. In the latter case, the restraints may couple the two set of coordinates directly \cite{Axelsen_1998,Ries_2022}, or the restraints may be defined relative to a common reference frame \cite{rocklin2013separated,Baumann_2023}. Finally, in ATM the two copies of the coordinates are held separate \cite{Azimi_2022}.

@@

Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed. One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation . This problem is similar to the required restraints in ABFE Calculations (see \ref{ABFErest}). Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on Boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-DCDT approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other. 
However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets. \cite{jespers2019qligfep, rieder2022lev}

@@

There are multiple strategies to alleviate such sampling challenges in an alchemical calculation. A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}. Another approach, known as integration over parts,\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recent,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate. In the integration over parts approach, restraining potentials can be used to restrict the sampling to specific macrostates,\cite{khuttan2023taming}. However, in practice, restraining potentials are sometimes omitted when the states are known to be separated by large free energy barriers and do not interconvert within the simulation time scale, such as cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}. A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation. This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from states separated by large free energy barriers) via replica exchange protocols. More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}. Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda}, and optimized alchemical enhanced sampling (ACES)~\cite{lee2023aces}  and approaches that integrate well-tempered metadynamics~\cite{barducci2008well} with alchemical methods~\cite{zhou2025zooming, khuttan2024make}, to name a few.

@@

\subsubsection{Which $\lambda$-sampling scheme will work best for my problem?}
\label{sec:sampling_schemes}
Though all alchemical simulations must sample from multiple $\vec{\lambda}$ states, different approaches can be used to achieve this. Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes. The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}). This type of scheme is currently used for some AMBER TI calculations~\cite{song2019using} and for Sire as implemented in BioSimSpace~\cite{hedges2019biosimspace}. However, if these simulations can be run simultaneously with communication between them, a simple extension allows mixing between these replicas. In this approach, the simulation at each $\vec{\lambda}$ can undergo periodic exchanges with neighboring $\vec{\lambda}$ values. This form of replica exchange, called Hamiltonian replica exchange, is based on ideas developed from Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}. With the Metropolis-Hastings acceptance criterion for exchanges, the generated ensemble of all replicas still samples from the Boltzmann distribution for each replica. This approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensional,sugita1999replicaexchange, woods2003development, jiang2010free}. The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{B}. It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltonian,hritz2007optimization},
AMBER~\cite{Lee_JChemInfModel_2020_v60_p5595}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
A specific variant of Hamiltonian replica exchange is FEP/REST2~\cite{rest2, feprest} that involves ``heating'' a small portion of the protein-ligand system to enhance the sampling within the binding pocket. It has been promoted as a great way to enhance sampling and is employed by default in Schr\"{o}dinger's FEP+ package. However, it should be noted that several studies have questioned its reliability and reported that it degrades the quality of free energy predictions as compared to those based on standard MD simulations without any enhanced sampling~\cite{wan2020fep+, bhati2022}. Therefore, a blind application of this sampling scheme is not recommended and it must be used with caution if at all. A variant of this method that overcomes some of the reported issues with FEP/REST2 is called alchemical enhanced sampling (ACES)~\cite{lee2023aces}. ACES creates localized targeted "enhanced sampling states" that are connected to the real state end-points through HREMD.  The method leverages the dual topology framework to produce a concerted “counter-diffusion” of alchemical states that leads to minimal rearrangement of the environment along the alchemical path~\cite{York_ACSPhysChemAu_2023_v3_p478}. The computational overhead of the method is negligible relative to more conventional alchemical free energy simulations with HREMD.

@@

	\item \href{https://ambermd.org/}{AMBER}, including its GPU-accelerated free energy engine~\cite{Lee_JChemTheoryComput_2017_v13_p3077,Giese_JChemTheoryComput_2018_v14_p1564,Lee_JChemInfModel_2018_v58_p2043} that supports a wide range of alchemical free energy methods~\cite{Tsai_JChemTheoryComput_2023_v19_p640,lee2023aces,Zhang_JChemTheoryComput_2024_v20_p3935}, workflows~\cite{Ganguly_JChemInfModel_2022_v62_p6069} and analysis tools~\cite{Giese_JChemInfModel_2025_v65_p5273,Case_JChemInfModel_2023_v63_p6183}. 
